\begin{center}
   SWE 619 \hfill Fall 2020 \\
   Final \hfill  Paul Ammann\\
\end{center}

This is an open-book/open-notes exam.
This means that you can access course materials on paper or on the internet.
The time limit on this exam is {\bf 2 hours and 30 minutes}.

It is a violation of the honor code to communicate with any
other person (except me, the instructor) about this exam
while you are taking it.

It is a violation of the honor code to
discuss or share the contents of this exam in any way
with any student who is currently registered for this course
but who has not yet completed this exam.

When you see a reference to code, you should
map that to the relevant Java file I provided as part of your study
guide for this exam. 

Capture your answers as a PDF document and submit on
Blackboard by 7:00PM.
If, for any reason, you have a problem submitting to BB, 
submit your final on Piazza in a private post.  Your post
should also explain your problem.

Please make sure your {\bf NAME} is at the top of your submission.

Code for the examples used in this exam is published on
guide discussed in class.  See the course web page for the relevant link.
Make sure you are getting the right code!

\vspace{0.5cm}

{\Large
\begin{center}
\begin{tabular}{|l|c|c|} \hline
Section		&	Points	&	Score \\ \hline \hline
Question 1	&	 30	& \\ \hline
Question 2	&	 35	& \\ \hline
Question 3 	&	 35	& \\ \hline
\hline
Total		&	100	& \\ \hline
\end{tabular}
\end{center}

%\vspace{1.0cm}
%\begin{center}
%{\large {\bf Name: \rule{4.5in}{0.01in}}}
%\end{center}
}
